---
title: sclanternR vignette
author: Jack Kamm
date: "Revised: 2026-08-21"
output:
  BiocStyle::html_document
package: sclanternR
vignette: >
  %\VignetteIndexEntry{sclanternR vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Example dataset

```{r load-packages, message = FALSE}
library(magrittr)
library(ggplot2)

library(ggtree)

library(sclanternR)
```

We illustrate sclanternR on a subset of the HAP1 cell-line data from
our preprint, consisting of the top 2000 loci from the full
analysis in our preprint.

We start with a matrix of allele counts in each locus in each cell:

```{r}
dim(hap1_subset)

hap1_subset[1:5,1:5]
```

The rows of this matrix are are locus-allele pairs, and the columns
are cells.  The rownames should follow the format `CHROM_POS_ALLELE`,
e.g. `chr1_100084142_TAAA`.

# Construct cell metadata

For our example dataset, the cells (column) names contain the ground truth clone and the cell barcode. We extract this metadata here so we can make plots with it later.

```{r}
cell_meta <- stringr::str_match(colnames(hap1_subset), "(.*)_(.*)")
colnames(cell_meta) <- c("cell_id", "clone", "cell_barcode")
cell_meta %<>% as.data.frame()
rownames(cell_meta) <- cell_meta$cell_id

head(cell_meta)

clone_colors <- pals::brewer.paired(12)[c(7,5,6,1,2)]
names(clone_colors) <- sort(unique(cell_meta$clone))
```

# Data object construction

We start by converting this into a `CellAlleleCounts` object:

```{r}
cac <- CellAlleleCounts(hap1_subset)
cac
```

Alternatively, if using PacBio long-read data processed by our
[sclantern-nf](https://github.com/Genentech/sclantern-nf) pipeline,
the `CellAlleleCounts` can be created by calling
`CellAlleleCounts_from_dataframe()` on the long table of counts
produced by that pipeline.

# Data slicing

Loci can be sliced along rows and cells can be sliced along columns:
```{r}
cac[1:5, 1:5]
```

It can also be alternatively treated like a list; the `i`th element
is a matrix of the counts at the `i`th locus:

```{r}
head(names(cac))

first <- cac[[1]] # same as cac[["chr1_100084142"]]

dim(first)

first[, 1:5]
```

# Run analysis pipeline with default parameters

The sclanternR procedure involves several steps, consisting of an initial clustering, selection of informative loci, smoothing of allele frequencies, and dimension reduction. The function `fitCellDistancePipeline()` runs each of these steps with automatic defaults.

```{r}
fit <- fitCellDistancePipeline(
    cac, verbose=1
)
```

## Examine PCA

We plot the PCA embedding produced by the pipeline.

```{r}
# shuffle order of cells for plotting
set.seed(12345)
shuffle_order <- sample(1:ncol(cac))

ggplot(
    cbind(
        cell_meta,
        fit$cell_pca
    )[shuffle_order,],
    aes(x=PC1, y=PC2, color=clone)
) +
    geom_point(shape=1) +
    scale_color_manual(values=clone_colors) +
    theme_bw(base_size=16)
```

## Tree building

For building distance-based phylogenies (e.g. with neighbor-joining), we recommend to use the *squared* Euclidean distances in the PCA space. These are also stored in the `dist` field of the fitted object.

```{r}
tr <- ape::nj(fit$dist)

ggtree(tr, layout='ape') %<+%
    cell_meta +
    ggrastr::rasterise(geom_point(
        data = function(d) {
            d %>% dplyr::filter(isTip) %>% .[shuffle_order,]
        },
        #shape=1,
        aes(color = clone)
    ), dpi=300) +
    scale_color_manual(values=clone_colors) +
    theme(legend.position='bottom')
```
