---
title: sclanternR vignette
author: Jack Kamm
date: "Revised: 2026-08-21"
output: rmarkdown::html_vignette
package: sclanternR
vignette: >
  %\VignetteIndexEntry{sclanternR vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Example dataset

```{r load-packages, message = FALSE}
library(magrittr)
library(ggplot2)

library(ggtree)

library(sclanternR)
```

We illustrate sclanternR on a subset of the HAP1 cell-line data from
our preprint, consisting of the top 2000 loci from the full
analysis in our preprint.

We start with a matrix of allele counts in each locus in each cell:

```{r}
dim(hap1_subset)

hap1_subset[1:5,1:5]
```

The rows of this matrix are are locus-allele pairs, and the columns
are cells.  The rownames should follow the format `CHROM_POS_ALLELE`,
e.g. `chr1_100084142_TAAA`.

# Construct cell metadata

For our example dataset, the cells (column) names contain the ground truth clone and the cell barcode. We extract this metadata here so we can make plots with it later.

```{r}
cell_meta <- stringr::str_match(colnames(hap1_subset), "(.*)_(.*)")
colnames(cell_meta) <- c("cell_id", "clone", "cell_barcode")
cell_meta %<>% as.data.frame()
rownames(cell_meta) <- cell_meta$cell_id

head(cell_meta)

clone_colors <- pals::brewer.paired(12)[c(7,5,6,1,2)]
names(clone_colors) <- sort(unique(cell_meta$clone))
```

# Data object construction

We start by converting the count matrix into a `CellAlleleCounts` object:

```{r}
cac <- CellAlleleCounts(hap1_subset)
cac
```

Alternatively, if using PacBio long-read data processed by our
[sclantern-nf](https://github.com/Genentech/sclantern-nf) pipeline,
the `CellAlleleCounts` can be created by calling
`CellAlleleCounts_from_dataframe()` on the long table of counts
produced by that pipeline.

# Data slicing

Loci can be sliced along rows and cells can be sliced along columns:
```{r}
cac[1:5, 1:5]
```

It can also be alternatively treated like a list; the `i`th element
is a matrix of the counts at the `i`th locus:

```{r}
head(names(cac))

first <- cac[[1]] # same as cac[["chr1_100084142"]]

dim(first)

first[, 1:5]
```

# Run analysis pipeline with default parameters

The sclanternR procedure involves several steps, consisting of an initial clustering, selection of informative loci, smoothing of allele frequencies, and dimension reduction. The function `fitCellDistancePipeline()` runs each of these steps with automatic defaults.

```{r}
fit <- fitCellDistancePipeline(
    cac, verbose=1
)
```

## Examine PCA

We plot the PCA embedding produced by the pipeline.

```{r, fig.width=5, fig.height=5}
# shuffle order of cells for plotting
set.seed(12345)
shuffle_order <- sample(1:ncol(cac))

ggplot(
    cbind(
        cell_meta,
        fit$cell_pca
    )[shuffle_order,],
    aes(x=PC1, y=PC2, color=clone)
) +
    geom_point(shape=1) +
    scale_color_manual(values=clone_colors) +
    theme_bw(base_size=16)
```

## Tree building

For building distance-based phylogenies (e.g. with neighbor-joining), we recommend to use the *squared* Euclidean distances in the PCA space. These are also stored in the `dist` field of the fitted object.

```{r, fig.width=5, fig.height=5}
tr <- ape::nj(fit$dist)

ggtree(tr, layout='ape') %<+%
    cell_meta +
    ggrastr::rasterise(geom_point(
        data = function(d) {
            d %>% dplyr::filter(isTip) %>% .[shuffle_order,]
        },
        #shape=1,
        aes(color = clone)
    ), dpi=300) +
    scale_color_manual(values=clone_colors) +
    theme(legend.position='bottom')
```

# Rerunning with better parameters

`fitCellDistancePipeline()` has several tuning parameters, and it tries to set reasonable defaults automatically. However, the automatic heuristics are imperfect, and it's recommended to visualize these cutoffs and set them manually. 

## Number of loci

The first cutoff to consider is the number of informative loci to retain. We use a scaled version of the mutual information (G-statistic) to rank the loci.

For the automatic cutoff, we test for the loci being independent of the clusters, and select loci at 5% FDR, with bounds of at least 10 loci or at most 300 loci selected.

Below, we plot the scaled mutual information scores and the cutoff selected. The maximum cutoff of 300 was chosen; however this was likely too high, as we can visually see a sharp drop at between 10-30 loci. Also, as a rule of thumb, we've found that selecting between 30-100 loci tends to give the best results.

```{r, fig.width=5, fig.height=5}
plot(fit$mi_cutoff)
```

## Number of principal components

The second cutoff to consider is the number of PCs to keep.

By default, we use the Gavish-Donoho procedure to select the number of PCs. But it's recommended to check the scree plot to see if there is a more obvious inflection point visually. In this case, there wasn't a more obvious cutoff, and we stick with the default selected 5 PCs.

```{r, fig.width=5, fig.height=5}
plot(fit$scree)
```

## Rerunning the pipeline

The pipeline can be rerun with the manually selected cutoffs. We can pass the previous fit in order to skip intermediate steps that are the same between the 2 runs (namely, the initial clustering procedure).

```{r}
fit2 <- fitCellDistancePipeline(
    cac,
    n_loci = 30,
    n_pcs = 5,
    init = fit,
    verbose=1
)
```

## Plot updated PCA

We plot the updated PCA from the rerun fit.

```{r, fig.width=5, fig.height=5}
ggplot(
    cbind(
        cell_meta,
        fit2$cell_pca
    )[shuffle_order,],
    aes(x=PC1, y=PC2, color=clone)
) +
    geom_point(shape=1) +
    scale_color_manual(values=clone_colors) +
    theme_bw(base_size=16)
```

## Plot updated tree

We recompute the neighbor joining tree and plot it.

```{r, fig.width=5, fig.height=5}
tr <- ape::nj(fit2$dist)

ggtree(tr, layout='ape') %<+%
    cell_meta +
    ggrastr::rasterise(geom_point(
        data = function(d) {
            d %>% dplyr::filter(isTip) %>% .[shuffle_order,]
        },
        #shape=1,
        aes(color = clone)
    ), dpi=300) +
    scale_color_manual(values=clone_colors) +
    theme(legend.position='bottom')
```

# Plotting heatmap of informative loci

It is also useful to plot the estimated cell-level allele frequencies
at the selected loci. The fit object contains these frequencies in the
`freq_list` field, which can be plotted with `heatmapLocusMatrixList()`.

```{r, fig.width=9, fig.height=8}
heatmapLocusMatrixList(
    fit2$freq_list,
    cell_meta[, 'clone', drop=FALSE],
    cell_meta_colors=list(
        clone=clone_colors
    )
)
```

By default the cells are sorted by a hierarchical clustering, but the
order can also be manually specified by setting `cluster_rows=FALSE`.
Below, we sort the cells by ground truth clone.

```{r, fig.width=9, fig.height=8}
o <- order(cell_meta$clone)

heatmapLocusMatrixList(
    lapply(fit2$freq_list, function(x) x[,o]),
    cell_meta[, 'clone', drop=FALSE],
    cell_meta_colors=list(
        clone=clone_colors
    ),
    cluster_rows = FALSE
)
```

`heatmapLocusMatrixList()` can also be used to plot other list of
matrices. Below, we demonstrate plotting the log10-scaled counts:

```{r, fig.width=9, fig.height=8}
log_cnt_list <- lapply(cac[names(fit2$freq_list)],
                       function(x) log10(as.matrix(x) + 1))

heatmapLocusMatrixList(
    log_cnt_list,
    cell_meta[, 'clone', drop=FALSE],
    cell_meta_colors=list(
        clone=clone_colors
    ),
    cluster_rows = FALSE,
    colors = rev(pals::cubehelix(10))
)
```
